---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "[YOUR NAME]"
date: "[DATE]"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# you could optionally set include = TRUE globally while working on the exercise and then just change it to include = FALSE before you hand in
# knitr::opts_chunk$set(include = FALSE)
```

# Template for the hand-in
### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:
    
- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?

### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries, include = FALSE}

library(pacman)
p_load(tidyverse, lme4)

```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

If your're in a project, just put the data in the project folder and you're good to go! (Or make a data subfolder to keep it tidy around here)
```{r Load Data, include = FALSE}

getwd()

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}
#loading data
data <- read.csv("language_dev_asd_clean.csv")

#right format 
str(data)

#conversion
data$SUBJ <- as.factor(data$SUBJ)
data$VISIT <- as.factor(data$VISIT)

#deleting records that are obviously fucked. 
data <- data[rowSums(is.na(data[ , 8:17])) < 10,] #2 observations bitch without any linguistic inf.

#well matched? 
summary(data$Diagnosis) #numerically, yes. (ASD: 176, TD: 196)

#creating subsets for age 
data_ASD <- data %>%
  filter(!is.na(Age) & VISIT == 1 & Diagnosis == "ASD")

data_TD <- data %>%
  filter(!is.na(Age) & VISIT == 1 & Diagnosis == "TD") 

#z-score for ASD
data_ASD <- data_ASD %>% 
  mutate(ageZ = scale(Age))

data_TD <- data_TD %>%
  mutate(ageZ = scale(Age))

#summary ASD
summary(data_ASD$ageZ) #bigger z-scores, some very old and young 
which(data_ASD$ageZ > 2) # 
which(data_ASD$ageZ < -2) #9

#summary TD
summary(data_TD$ageZ) #somewhat smaller z-scores 
which(data_TD$ageZ > 2) #31 
which(data_TD$ageZ < -2) #none

#comparing means 
difference <- mean(data_TD$Age) - mean(data_ASD$Age)
difference #very small difference. 

#histograms
hist(data_ASD$ageZ) 
hist(data_TD$ageZ)

hist(data_ASD$Age)
hist(data_TD$Age) #don't get quite as old - matching production. 

#Gender
summary(data_ASD$Gender) #Female = 146, Male = 26
summary(data_TD$Gender) #Female = 155, Male = 35

#Ethnicity
summary(data_ASD$Ethnicity) #ASD mainly white & some scattered. 
summary(data_TD$Ethnicity) #TD only white & asian 

#random intercepts
data$Kid = factor(data$SUBJ) #from lecture 

```

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}

#plotting
ggplot(data, aes(as.numeric(VISIT), CHI_MLU, color = Diagnosis)) +
  geom_point() + 
  geom_smooth(method = "lm")

ggplot(data, aes(as.numeric(VISIT), CHI_MLU)) + 
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~ Diagnosis)

ggplot(data, aes(Diagnosis, CHI_MLU))+
  geom_boxplot() #outliers in ASD. 

#linear mixed effects model 
library(lme4)

model_data <- data %>%
  filter(!is.na(CHI_MLU) & !is.na(VISIT) & !is.na(Diagnosis))

#is diagnosis a significant predictor of MLU? - no
mixed_model <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)
summary(mixed_model)

model_simple <- lmer(CHI_MLU ~ VISIT + (1+VISIT|SUBJ), model_data, REML=FALSE)
anova(model_simple, mixed_model) #not significantly different with diagnosis

#is visit a significant predictor of MLU? - yes
model_simple <- lmer(CHI_MLU ~ Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)
anova(model_simple, mixed_model) #very significant. 

```

How would you evaluate whether the model is a good model?

```{r ex2 evaluate, include = FALSE}
mixed_model <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)
mixed_model

```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better

```{r ex2 growth curve, include = FALSE}

#quadratic 
ggplot(data, aes(as.numeric(VISIT), CHI_MLU, color = Diagnosis)) +
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ poly(x, 2))

#Cubic
ggplot(data, aes(as.numeric(VISIT), CHI_MLU, color = Diagnosis)) +
  geom_point() + 
  stat_smooth(method = "lm", formula = y ~ poly(x, 3))

#different models 
m_linear <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)
m_quadratic <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)
m_cubic <- lmer(CHI_MLU ~ VISIT + Diagnosis + (1+VISIT|SUBJ), data = model_data, REML=FALSE)

```

Exciting right?
Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your best model's predictions

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... 

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}

```

[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Malte (au540041@post.au.dk).


```{r ex4, include = FALSE}

```

[REPORT THE RESULTS]

